---
title: "Embedding Gower's metric in $\\mathbb{R}^k/L_1$"
author: "Kirill Müller"
date: "`r Sys.Date()`"
output: rmarkdown::html_document
bibliography: mangow.bib
abstract: >
  This draft describes an embedding of Gower's dissimilarity index into the
  $L_1$ metric on $\mathbb{R}^k$, where $k$ depends on the structure of the variables,
  and only in some cases on the data.
  This embedding allows using standard algorithms such as the kd-tree
  to find the nearest neigbors of an observation without processing all other
  observations in the dataset,
  which can be useful in applications such as clustering or statistical matching.
  An implementation for the statistical software package R is demonstrated.
vignette: >
  %\VignetteIndexEntry{Embedding Gower's metric in $\\mathbb{R}^k/L_1$}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


# Definition

Gower's dissimilarity index `r knitcitations::citep("10.2307/2528823")`
is a distance measure for mixed-type multivariate data.
The distance $d(x, y)$ between two tuples $x = (x_1, x_2, \ldots x_n)$
and $y = (y_1, y_2, \ldots y_n)$
is a weighted sum of the distances for each variable $x_i$ and $y_i$;
the weights $w_i$ are chosen so that they are nonnegative and sum up to 1:

$$
d(x,y) = \sum_{i=1}^n w_i d(x_i, y_i)
$$

For individual variables,
the calculation of the distance $d(x_i, y_i)$ depends on the type of the variable:

- For interval-scaled variables, the range is mapped to $[0, 1]$:
  $d(x_i, y_i) = |f(x_i) - f(y_i)|$ with
  $f(z) := \frac{z - z_{\text{min}}}{z_{\text{max}} - z_{\text{min}}}$.

- Ordinal variables are treated like interval-scaled variables.

- For nominal variables, the distance is 0 if and only if the values are equal,
  otherwise it is 1:
  $d(x_i, y_i) = 1 - I(x_i = y_i)$.

- Binary variables are outside the scope of this discussion
  but will be considered in the conclusion.

The variable types correspond to R's `numeric`, `factor`, and `ordered`,
and `logical` types, respectively.


# Embedding univariate data

`r knitcitations::citet("Gower_1966")` discusses the concept of a Q matrix
where each of $n$ observations is mapped to an $n$-dimensional vector,
and dimensionality reduction techniques are applied.

For embedding Gower's distance into $L_1$,
we first consider univariate data,
and define a transformation $f$ to $\mathbb{R}^k$ for some $k$ so that
$$
||f(x_i) - f(y_i)||_1 = d(x_i, y_i)
$$

The function $f$ from the definition already yields an embedding into $L_1$,
mapping to $\mathbb{R}^1$ for univariate interval-scaled and ordinal data.
The embedding for nominal variables is a bit more difficult.
For $x_i \in { a_1, a_2, \ldots a_m}$,
the values are mapped to $\mathbb{R}^{\left\lceil \frac{m}{2} \right\rceil}$ as follows:

- $f(a_1) := (\frac{1}{2}, 0, 0, \ldots)$
- $f(a_2) := (-\frac{1}{2}, 0, 0, \ldots)$
- $f(a_3) := (0, \frac{1}{2}, 0, \ldots)$
- $f(a_4) := (0, -\frac{1}{2}, 0, \ldots)$
- $f(a_5) := (0, 0, \frac{1}{2}, \ldots)$
- $\ldots$

The $L_1$ distance between any pair of different values is 1:

- $||f(a_{2k}) - f(a_{2k + 1})||_1 = |0| + \cdots + \left|\frac{1}{2} - \left(- \frac{1}{2}\right)\right| + \cdots + |0| = 1 = d(a_{2k}, a_{2k + 1})$

- $||f(a_{2k}) - f(a_{2k + j})||_1 = |0| + \cdots + \left|\frac{1}{2}\right| + \cdots + \left|\pm \frac{1}{2}\right| + \cdots + |0| = 1 = d(a_{2k}, a_{2k + j})$ for $j \geq 2$

- $||f(a_{2k + 1}) - f(a_{2k + j})||_1 = |0| + \cdots + \left|-\frac{1}{2}\right| + \cdots + \left|\pm \frac{1}{2}\right| + \cdots + |0| = 1 = d(a_{2k + 1}, a_{2k + j})$ for $j \geq 2$

This defines an exact embedding into $L_1$ for single interval-scaled, ordinal,
and nominal variables:
Gower's dissimilarity for the original data equals the $L_1$ distance for the transformed data,
which is a single real (for interval-scaled and ordinal variables) or a tuple of reals (for nominal variables).


# Embedding mixed-type multivariate data

The embedding of multivariate data consists of scaling and concatenating the $L_1$ embedding of the individual variables to tuples.
Assuming that $x_1$ is embedded into $\mathbb{R}^{m_1}$, $x_2$ into $\mathbb{R}^{m_2}$ etc.,
the tuple $(x_1, x_2, \ldots, x_n)$ can be embedded in $\mathbb{R}^{m_1 + m_2 + \ldots m_n}$
by defining

$$
f(x_1, x_2, \ldots, x_n) = w_1 f(x_1) \oplus w_2 f(x_2) \oplus \cdots \oplus w_n f(x_n)
$$

with $\oplus$ defined as "vector concatenation".

described in the previous section

The generalization to multiple variables simply applies the weights to the transformed data and binds them together.

Example:


# Examples


```{r}
mangow:::mangow_one(1:5, "interval")
mangow:::mangow_one(ordered(letters[1:5]), "ordinal")
mangow:::mangow_one(factor(LETTERS[1:5]), "nominal")
```

(Weighting is currently not supported, but can be added in a straightforward way.)

```{r}
mangow::mangow(data.frame(interval = 1:5,
                          ordinal = ordered(letters[1:5]),
                          nominal = factor(LETTERS[1:5])))
```


# Conclusion

- Binary variables


# Bibliography

```{r echo=FALSE, results='asis'}
knitcitations::bibliography()
```
